---
title: "Metabolite_survival: lipids"
author: "Zeyuan Song & Paola Sebastiani"
date: "June 2023"
output:
  html_document:
    theme: united
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapse: false
---

# Initialize paths
```{r}
setwd("where you work")
dir() 
llfs.pheno.dir <-  where phenotypic data is
llfs.metab4.dir <- where lipidomic data is
annot.dir <- where annotation file is
  
library(readxl)
library(dplyr)
library(ggplot2)
library(Heatplus)
library(tidyverse)
library(coxme)
library(GENESIS)
suppressPackageStartupMessages(library(SeqArray))
suppressPackageStartupMessages(library(SeqVarTools))
library(Biobase)
```
# Function
```{r }
 standardize.col <- function(data.in){
     data.out <- data.in
       for(j in 1: ncol(data.in)){
           data.out[,j] <- (data.in[,j]-mean(data.in[,j], na.rm=T))/sqrt(var(data.in[,j], na.rm=T))
       }
     data.out
 }
```
# Read data and basic QC
# Read LLFS list for batch 4
# use fake names of metabolites for analysis and annotate at the end
```{r }
llfs.data.batch4 <- read.csv(paste0(llfs.pheno.dir,  "pheno.data"), 
                             header=T, na.strings = c("", NA))
   dim(llfs.data.batch4)
    met.4.llfs <- readr::read_csv(paste0(llfs.metab4.dir, "lipids_normalized"))
     n.metab <- ncol(met.4.llfs)-2
     orig.metab.names <- names(met.4.llfs)[3:ncol(met.4.llfs)]
     fake.metab.names <- paste0("metab", c(1:n.metab))
     metab.look.up.table <- data.frame( orig.metab.names, fake.metab.names)
     names(met.4.llfs)[3:ncol(met.4.llfs)] <- fake.metab.names
       metab.look.up.table$orig.metab.names <- str_replace_all(metab.look.up.table$orig.metab.names,
                                                               "dCer_NS", "Cer")
```

# Data  formatting-- for now analyze data at visit 1

```{r }
met.llfs <- met.4.llfs
dim(met.llfs)
 table(met.llfs$visitcode)
 
met.llfs.vst1 <- met.llfs %>%
  filter(visitcode==1 | visitcode==4 | visitcode==7)
met.data.1 <- as.data.frame(t(met.llfs.vst1[ , 3:ncol(met.llfs.vst1)]))
dim(met.data.1)
colnames(met.data.1) <- met.llfs.vst1$subject
```

# Data visualization and outlier detection
```{r }
hist(apply(met.data.1,2,min))
summary(apply(met.data.1,2,min))
boxplot(log(met.data.1))

pca.res <- prcomp(log(t(met.data.1)), scale. = T)
  # summary(pca.res)
   plot(pca.res)
    plot(pca.res$x[,1:2])
      outliers <- names(which(pca.res$x[,1] < -25 & pca.res$x[,2] < 0))
      print(outliers)
      ok.samples <- setdiff(met.llfs.vst1$subject, outliers)
  new.met.data.1 <- met.data.1 %>%
    select(as.character(ok.samples))

  pca.res <- prcomp(log(t(new.met.data.1)), scale. = T)
   # summary(pca.res)
   plot(pca.res)
    plot(pca.res$x[,1:2])
  
# Drop outliers  
  met.llfs.vst1 <- met.llfs %>%
   filter(visitcode==1 | visitcode==4 | visitcode==7) 
    met.llfs.vst1 <- met.llfs.vst1[match(as.character(ok.samples), met.llfs.vst1$subject),]
  
    # now drop outliers
    met.data.1 <- as.data.frame(met.llfs.vst1[ ,3:ncol(met.llfs.vst1)])
    for(ind.col in 1:ncol(met.data.1)){
          this.metab <- log(met.data.1[,ind.col])
           this.mean <- mean(this.metab, na.rm=T)
           this.var  <- var(this.metab, na.rm=T)
              set.to.na <- which((this.metab > this.mean+4*sqrt(this.var)) | 
                               (this.metab < this.mean-4*sqrt(this.var)))
              met.data.1[set.to.na, ind.col ] <-  exp(this.mean)
    }
    
     met.llfs.vst1 <- data.frame(met.llfs.vst1[,1:2], met.data.1)

met.data.1 <- as.data.frame(t(met.llfs.vst1[ , 3:ncol(met.llfs.vst1)]))
dim(met.data.1)
colnames(met.data.1) <- met.llfs.vst1$subject

my.annot <-names(met.data.1)
   my.dat <- as.matrix(t(met.data.1)); dim(my.dat) <- dim(t(met.data.1));  my.dat <- t(my.dat)
   my.map <- annHeatmap2(log(my.dat),ann=list(my.annot))
plot(my.map)

```
No evident bias -- good normalization

# analysis data
we will use cohph with cluster by family 
```{r}
llfs.master.vst1 <- llfs.data.batch4
 analysis.master.dat <- data.frame(met.llfs.vst1, llfs.master.vst1[match(met.llfs.vst1$subject, llfs.master.vst1$subject),])
 dim(analysis.master.dat)
 
 length(unique(analysis.master.dat$subject))
   hist(analysis.master.dat$BirthYear)
```

# pca 
```{r}
pc.df <- read.csv(PCA data file here)

grm <- readRDS(GRM file here)

analysis.dat <- left_join(analysis.master.dat, pc.df, by=c("subject"="sample.id"))

ggplot(analysis.dat, aes(PC1, PC2)) + geom_point()
ggplot(analysis.dat, aes(PC3, PC4)) + geom_point()
```
# association analysis
# data preparation
```{r}
  analysis.dat$Status <- 0; analysis.dat$Status[which(analysis.dat$Alive == "No")] <-1

analysis.final.dat <- analysis.dat %>%
  mutate(FC_DK = FC == "DK") %>%
   mutate( BYC = BirthYear < 1935) %>%
    mutate( FU = Age-Age.e) %>%
  select(subject, pedid, Status, FU, Age, Age.e, Sex, Education, BYC, FC_DK, PC1, PC2, PC3, PC4, htn_meds, lipid_meds, nitro_meds,
         t2d_meds, rownames(met.data.1))
dim(analysis.final.dat)
 

 #### use family ID as clustering variable
      analysis.final.dat <- analysis.final.dat[ order(analysis.final.dat$pedid),]
        unique.id <- unique(analysis.final.dat$pedid) 
     analysis.final.dat$new.fam.id <- rep(NA, nrow(analysis.final.dat))
     for(i in 1:length(unique.id)){
          analysis.final.dat$new.fam.id[ which(analysis.final.dat$pedid == unique.id[i])] <- i}
     # check that families are stucked
     plot(analysis.final.dat$new.fam.id)
     
var.list <- c("new.fam.id","BYC","Status","FU", "Age", "Age.e", "Education","Sex","htn_meds","lipid_meds","nitro_meds",
              "t2d_meds","FC_DK","PC1","PC2","PC3","PC4")
skimr::skim(analysis.final.dat[,var.list])

analysis.final.no.missing.dat <- na.omit(analysis.final.dat)
dim(analysis.final.no.missing.dat)
 

analysis.final.no.missing.dat[,rownames(met.data.1)] <- standardize.col( log(analysis.final.no.missing.dat[,rownames(met.data.1)]) )
 write.table(analysis.final.no.missing.dat, "analysis.final.no.missing.dat.batch4.csv", sep=",", row.names=F)
```

## GEE analysis, stratified by BYC
# run analysis
```{r }
covariates <- c("Age.e", "Education","Sex","FC_DK","PC1","PC2","PC3","PC4","htn_meds","lipid_meds","nitro_meds","t2d_meds")
out_dat <- c()
for(i.met in rownames(met.data.1)){
  fix.eff=paste("Surv(FU, Status) ~",i.met)
    if (!is.null(covariates)) {for (covi in covariates) fix.eff=paste(fix.eff,"+",covi) }
    fix.eff <- paste(fix.eff, "+cluster( new.fam.id)")
    fix.eff <- formula(fix.eff)
   fit <- try(coxph( fix.eff,data=analysis.final.no.missing.dat))
   
   coeff <- coefficients(summary(fit))
  out_dat <- rbind(out_dat, data.frame(metabolite = i.met, n_obs = summary(fit)$nevent,
                     metab_eff = coeff[1,1], 
                     metab_HR = coeff[1,2],
                     metab_sd = coeff[1,4], 
                     metab_pval = coeff[1,6], 
                    AgeE_eff = coeff["Age.e",1], 
                    AgeE_sd = coeff["Age.e",4], 
                    AgeE_pval = coeff["Age.e",6], 
                       Male_eff = coeff["SexMale",1], 
                       Male_sd = coeff["SexMale",4], 
                       Male_pval = coeff["SexMale",6],
                     Educ_eff = coeff["Education",1], 
                     Educ_sd = coeff["Education",4], 
                     Educ_pval = coeff["Education",6],
                       FC.DK_eff = coeff["FC_DKTRUE",1], 
                       FC.DK_sd = coeff["FC_DKTRUE",4], 
                       FC.DK_pval = coeff["FC_DKTRUE",6],
                     PC1_pval = coeff["PC1",6], 
                     PC2_pval = coeff["PC2",6], 
                     PC3_pval = coeff["PC3",6], 
                     PC4_pval = coeff["PC4",6],
                       htn_med_eff = coeff["htn_meds",1], 
                       htn_med_pval = coeff["htn_meds",6],
                       lipid_med_eff = coeff["lipid_meds",1], 
                       lipid_med_pval = coeff["lipid_meds",6],
                      nitro_med_eff = coeff["nitro_meds",1], 
                      nitro_med_pval = coeff["nitro_meds",6],
                        t2d_med_eff = coeff["t2d_meds",1], 
                        t2d_med_pval = coeff["t2d_meds",6]))
}

out_dat1 <- as.data.frame(out_dat) %>%
   mutate(Metab_pval_adj = p.adjust(out_dat$metab_pval, method="BH")) %>%
   arrange(metab_pval)
 nsign <- length(which(out_dat1$Metab_pval_adj < 0.05))
   
   out_dat2 <- left_join(metab.look.up.table, out_dat1, by=c("fake.metab.names" = "metabolite"))
write.csv(out_dat2, "Survival_stand.gee_batch4.pc_genesis.03.30.2023.csv", row.names=F)
```

#Annotation
```{r}
annot.dat <- read.csv(paste0(annot.dir, "llfs.annotation.03.30.2023.csv")) %>%
  mutate(Compound.Name = Input.name)

sum(out_dat2$metabolite %in% annot.dat$Compound.Name)
#188

out_dat1_annot <- data.frame(annot.dat, out_dat2[match(annot.dat$Compound.Name, out_dat2$orig.metab.names),]) %>%
  arrange(metab_pval)
write.csv(out_dat1_annot, "annotated_Survival_stand.gee_batch4.pc_genesis.03.30.2023.csv")
```

# Plot
```{r}
out_dat1_annot <- read.csv( "annotated_Survival_stand.gee_batch4.pc_genesis.03.30.2023.csv", header=T)

out_dat1_annot <- out_dat1_annot[order(out_dat1_annot$metab_pval), ]
 nsign <- length(which(out_dat1_annot %>% pull(Metab_pval_adj) < 0.05))

 analysis.final.no.missing.dat <- read.csv("analysis.final.no.missing.dat.batch4.csv", header=T)
require(survminer)
metab <- as.character(out_dat1_annot$fake.metab.names[1:nsign])
metab.names <- str_replace(as.character(out_dat1_annot$Compound.Name[1:nsign]), "/"," ")

 j <-0
for(i in metab){
   j <- j+1
   analysis.final.no.missing.dat$group = "L/Y"
    analysis.final.no.missing.dat$group[which(analysis.final.no.missing.dat[,i]>
                                    mean(analysis.final.no.missing.dat[,i]) & analysis.final.no.missing.dat$BYC == FALSE)] <-"H/Y"
    analysis.final.no.missing.dat$group[which(analysis.final.no.missing.dat[,i]>
                                    mean(analysis.final.no.missing.dat[,i]) & analysis.final.no.missing.dat$BYC == TRUE)] <-"H/O"
   analysis.final.no.missing.dat$group[which(analysis.final.no.missing.dat[,i]<=
                                    mean(analysis.final.no.missing.dat[,i]) & analysis.final.no.missing.dat$BYC == TRUE)] <-"L/O"
  fit <- survfit(Surv(FU, Status) ~ group, data = analysis.final.no.missing.dat) 
    p<-ggsurvplot(fit, data = analysis.final.no.missing.dat,
               xlim=c(0,15), xlab="FU_time", 
          font.tickslab = c(25, "bold", "#993333"),
          font.x =c(25, "bold", "black"), 
          font.y =c(25, "bold", "black"),
          font.legend =c(20, "bold", "black"),
          font.main =c(25, "bold", "black"),
          main=metab.names[j],
           margin = margin(0,0,-20,0))+
          guides(color=guide_legend(nrow=2,byrow=TRUE))+
      labs(title=metab.names[j])
  png(paste0("plot_dir/", metab.names[j], "05.2023.png"), width = 480, height = 480)
  print(p)
  dev.off()
}
```

